\chapter*{Acknowledgements}
\thispagestyle{empty}

I would like to express my sincere gratitude to all those who have supported me throughout the completion of this thesis. My deepest thanks go to my supervisors, Khaled Wahba and Dr. Sayantan Auddy, for their invaluable guidance, insightful feedback, and weekly encouragement. I am also grateful to Prof. Dr. Wolfgang Hönig and the other colleagues in the research group for their constructive discussions and suggestions.

I would additionally like to acknowledge the foundational work of the researchers who preceded me, and to highlight the main external libraries and tools that were instrumental in this work:

\begin{itemize}
  \item \texttt{MJX (MuJoCo \cite{todorov2012mujoco})} - a high-performance physics engine designed for model-based optimization and control in robotics and biomechanics, providing accurate rigid-body simulation with contacts and constraints.
  \item \texttt{JAXMARL} \cite{flair2023jaxmarl}  - a library built on JAX for scalable multi-agent reinforcement learning, offering modular implementations of algorithms and utilities for fast experimentation with vectorized environments.
  \item \texttt{JAX} \cite{jax2018github} - Numerical computing library that combines composable function transformations with NumPy-compatible APIs for high-speed, highly GPU parallelized reinforcement learning research.
  \item \texttt{ChatGPT (o4 model)} - OpenAI's large language model used throughout this thesis for drafting and refining text, generating code snippets, and conducting exploratory research.
  \item \texttt{Grammarly} - an AI-powered writing assistant that provides real-time grammar, spelling, and style suggestions to improve clarity and readability of the thesis. Used for proofreading and enhancing the overall quality of the text.
\end{itemize}

\vspace{1em}
\noindent
Without the support of these individuals and the capabilities provided by these tools, this research would not have been possible.
