\chapter*{Declaration}
\thispagestyle{empty}
\subsubsection{English}
I hereby declare that I have composed this thesis independently, without the assistance of third parties, and solely using the sources and tools listed. All passages taken from these sources and tools have been duly identified as such.
In cases where generative AI tools were employed, I have specified the product names, manufacturers, the exact software versions used, and their respective purposes. I assume full responsibility for the selection, incorporation, and all outcomes of the AI-generated content utilized.
In the development of this thesis, I employed OpenAI ChatGPT (model o4) for coding and research assistance. Additionally, I leveraged ChatGPT as well as Grammarly to review and enhance writing, ensuring correctness in grammar and overall readability. \todo{clarify chatgpt declaration}
\subsubsection{German}
Hiermit versichere ich, dass ich die vorliegende Arbeit eigenständig ohne Hilfe Dritter und ausschließlich unter Verwendung der aufgeführten Quellen und Hilfsmittel angefertigt habe. Alle Stellen die den benutzten Quellen und Hilfsmitteln unverändert oder sinngemäß entnommen sind, habe ich als solche kenntlich gemacht.
Sofern generative KI-Tools verwendet wurden, habe ich Produktnamen, Hersteller, die jeweils verwendete Softwareversion und die jeweiligen Einsatzzwecke (z.B. sprachliche Überprüfung und Verbesserung der Texte, systematische Recherche) benannt. Ich verantworte die Auswahl, die Übernahme und sämtliche Ergebnisse des von mir verwendeten KI-generierten Outputs vollumfänglich selbst.
\vspace{3cm}
\noindent{}Berlin, \\\today
\begin{minipage}[t]{8cm}
\centering \hspace{20mm} \hrulefill \\
\hspace{20mm}Viktor Lorentz
\end{minipage}